Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-07 Manuscript/manuscript.tex
Party updated intro, not ready yet

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{JOE: list more introductory examples, where it has interesting effects.}

Zwitterionic phosphocholine (PC) lipid bilayers are common model systems for
cellular membranes. Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$ are with 
have been widely used as
model systems to study details of cation interactions with cellular membranes
by using experimental \cite{cevc90,tocanne90,binder02,pabst07,uhrikova08}
and theoretical methods \cite{bockmann03,bockmann04,Berkowitz12,??}

While relative binding affinity of different ions is agreed to follow Hoffmeister
series, the molecular details of binding and binding energetics are
not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
\todo{PAVEL: introduce previous theoretical work that discusses cation binding to POPC w.r.t. its specific moiteties, e.g. Lukas' paper}.

%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nucelar magnetic resosnance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02}. 
Molecular level interpretation of the results, however, requires assumptions
about binding models and is often not fully conclusive \cite{??}.
On the other hand, the accuracy of the state of the art atomistic resolution lipid and ion models
have turned out insufficient for the detailed interpretation of cation binding details \cite{catte16}.
The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions (e.g.~divalent cation-induced charge inversion of bacterial membranes~\cite{Luan2016}
\todo{JOE: this may be a too daring statement. \\
  SAMULI: I think that the divalent cation induced charge inversion is not an artefact,
  but it might happen with too small concentrations. 
  On the other hand, it is not a new observation either, see e.g. literature cited in NMRlipids IV project.}).

@@

In this work we show that the accuracy of ion binding, especially calcium,
in zwitterionic PC lipid bilayer can be significantly improved 
for classical MD simulation models of zwitterionic lipids
by including electronic polarizability
in the polar region of lipid molecules
using the electronic continuum correction (ECC)~\cite{leontyev11}. 
This approach has been previously shown to improve performace of
ion models against neutron scattering data in concentrated solutions 
and ab-initio simulations~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}. 
The realistic structure of ions in a bulk solvent was not, however, sufficient to correct binding in lipid
bilayers~\cite{catte16}. 
To fix it, we propose
a classical fixed-charge MD simulation model of 
1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) lipid 
that accounts for electronic polarizability by using ECC. 
We will validate such a model by comparing its lipid bilayer structue
to x-ray scattering data and its head group structure to NMR experiments. 
We will then quantify the accuracy of the head group order parameter response to the bound charge
and compare it to the experimental NMR data of sodium and calcium binding in PC lipid bilayers. 
In addition, we will quantify relative binding affinities of different moieties in POPC to such cations. 
This will be discussed in terms of binding modes, stoichiometry and binding isotherms and their models used in experiments.
The proposed POPC model, ECC-lipids, and the approach to obtain it
is, hence, highly useful for future MD simulations with physiological salt conditions, 
and for effects where membrane electrostatics play a major role.

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
like in signal propagation in neurons and vesicle fusion.
\todo{This could be maybe extended.}
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple the model systems can be then
used to understand the role of ions in complex biological systems.

Zwitterionic phosphocholine (PC) lipid bilayers are commonly used model systems for
cellular membranes. Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied with
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not agreed in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in PC lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Interpretation of these experiments suggests that Na$^+$ ions have
negligible binding in PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules. Atomistic resolution
molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for both ions and interactions with 3-4 lipids,
including also interactions with carbonyl
oxygens~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.
Some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies by introducing a direct
method to compare ion binding affitity between simulations and
experiments by using the electrometer concept~\cite{seelig87}.
Using massive data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

%\todo{PAVEL: introduce previous theoretical work that discusses cation binding to POPC w.r.t. its specific moiteties, e.g. Lukas' paper}.

%While relative binding affinity of different ions in PC lipids is agreed to follow Hoffmeister
%series, the molecular details of binding and binding energetics are

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved by implicitly including electronic polarizability
in the polar region of lipid molecules. The electronic
polarizability is included by using the electronic
continuum correction (ECC)~\cite{leontyev11}, which
has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}.

We will validate such a model by comparing its lipid bilayer structue
to x-ray scattering data and its head group structure to NMR experiments. 
We will then quantify the accuracy of the head group order parameter response to the bound charge
and compare it to the experimental NMR data of sodium and calcium binding in PC lipid bilayers. 
In addition, we will quantify relative binding affinities of different moieties in POPC to such cations. 
This will be discussed in terms of binding modes, stoichiometry and binding isotherms and their models used in experiments.
The proposed POPC model, ECC-lipids, and the approach to obtain it
is, hence, highly useful for future MD simulations with physiological salt conditions, 
and for effects where membrane electrostatics play a major role.

The overestimated specific cation binding observed in current lipid models \cite{catte16}
may lead to articifially positively charged memranes and significant artefacts in
MD simualtions (e.g.~divalent cation-induced charge inversion of bacterial membranes~\cite{Luan2016}

%\todo{JOE: this may be a too daring statement. \\
%  SAMULI: I think that the divalent cation induced charge inversion is not an artefact,
%  but it might happen with too small concentrations. 
%  On the other hand, it is not a new observation either, see e.g. literature cited in NMRlipids IV project.}).
